---
title: "Integrating Spectra with Python's matchms package"
package: SpectriPy
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Integrating Spectra with Python's matchms package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Mass Spectrometry, MS, MSMS, Metabolomics, Infrastructure, Quantitative}
  %\VignettePackage{SpectriPy}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{Spectra,BiocStyle,SpectriPy,basilisk,reticulate}
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
BiocStyle::markdown()
```

**Compiled**: `r date()`

# Introduction

The *SpectriPy* package allows integration of Python MS packages into a
*Spectra*-based MS analysis in *R*. Python functionality is wrapped into R
functions allowing a seamless integration of the functionality of Python's
*matchms* package into R. In addition, functions to convert between R's
`Spectra` objects and Python's *matchms* spectrum objects are available to the
advanced user or developer enabling to create custom functions or workflows on
`Spectra` objects in Python and executing them in R using the *reticulate* R
package.


# Installation

The package requires a python environment to be available and can be installed
with the *BiocManager* R package using the command
`BiocManager::install("RforMassSpectrometry/SpectriPy")`. All required python
libraries are installed automatically on demand. Note that first installation
or first invocation of specific functions might take long because of this
installation process. The *BiocManager* package, if not already available, could
be installed with `install.packages("BiocManager")`.


# Spectra similarity calculations using `matchms`

The *SpectriPy* package provides the `compareSpectriPy()` function that allows
to perform spectra similarity calculations using the scoring functions from the
*matchms* Python package. Below all currently supported scoring functions are
listed along with the *parameter* class that allows selecting and configuring
the algorithm in the `compareSpectriPy()` function. Additional functions will be
added in future.

- [*CosineGreedy*](https://matchms.readthedocs.io/en/latest/api/matchms.similarity.CosineGreedy.html):
  `CosineGreedyParam`.
- [*CosineHungarian*](https://matchms.readthedocs.io/en/latest/api/matchms.similarity.CosineHungarian.html):
  `CosineHungarianParam`.
- [*ModifiedCosineParam*](https://matchms.readthedocs.io/en/latest/api/matchms.similarity.ModifiedCosine.html):
  `ModifiedCosineParam`.

We next create some simple example spectra and subsequently use the
`compareSpectriPy()` function to calculate pairwise similarities between these.

```{r, message = FALSE}
library(Spectra)
library(SpectriPy)

## Create a Spectra object with two MS2 spectra for Caffeine.
caf <- DataFrame(
    msLevel = c(2L, 2L),
    name = "Caffeine",
    precursorMz = c(195.0877, 195.0877)
)
caf$intensity <- list(
    c(340.0, 416, 2580, 412),
    c(388.0, 3270, 85, 54, 10111))
caf$mz <- list(
    c(135.0432, 138.0632, 163.0375, 195.0880),
    c(110.0710, 138.0655, 138.1057, 138.1742, 195.0864))
caf <- Spectra(caf)

## Create a Spectra object with two MS2 spectra for 1-Methylhistidine
mhd <- DataFrame(
    msLevel = c(2L, 2L),
    precursorMz = c(170.0924, 170.0924),
    id = c("HMDB0000001", "HMDB0000001"),
    name = c("1-Methylhistidine", "1-Methylhistidine"))
mhd$mz <- list(
    c(109.2, 124.2, 124.5, 170.16, 170.52),
    c(83.1, 96.12, 97.14, 109.14, 124.08, 125.1, 170.16))
mhd$intensity <- list(
    c(3.407, 47.494, 3.094, 100.0, 13.240),
    c(6.685, 4.381, 3.022, 16.708, 100.0, 4.565, 40.643))
mhd <- Spectra(mhd)
```

We first calculate pairwise similarities between all spectra defined above and
those of caffeine using *Spectra*'s built-in `compareSpectra()` function.

```{r}
all <- c(caf, mhd)
res_r <- compareSpectra(all, caf)
res_r
```

Thus, `compareSpectra()` returned the pairwise similarity scores (by default
calculated using the normalized dot-product function) between all spectra in
`all` (rows) and all spectra in `caf` (columns). `compareSpectriPy()` works
similar, with the difference that we need to specify and configure the
similarity function (from *matchms*) using a dedicated parameter object. Below
we calculate the similarity using the *CosineGreedy* function changing the
`tolerance` to a value of `0.05` (instead of the default `0.1`).

```{r}
res <- compareSpectriPy(all, caf, param = CosineGreedyParam(tolerance = 0.05))
res
```

As a result `compareSpectriPy()` returns also a numeric matrix of similarities.
Note also that the first `compareSpectriPy()` call takes usually a little longer
because the Python setup has to be initialized.

Next we use the *ModifiedCosine* algorithm that considers also differences
between the spectra's precursor *m/z* in the calculation.

```{r}
res <- compareSpectriPy(all, caf, param = ModifiedCosineParam())
res
```

A third example is the *CosineHungarian* algorithm, also a similarity function from *matchms*. This algorithm calculates the *cosine similarity score* as with `CosineGreedyParam`, but using the Hungarian algorithm to find the best matching peaks between the compared spectra. The algorithm can be configured with parameters `tolerance`, `mzPower` and `intensityPower` (see parameter description for more details).

```{r}
res <- compareSpectriPy(all, caf, param = CosineHungarianParam())
res
```

A fourth example is the *NeutralLossesCosine* algorithm, also a similarity function from *matchms*. The neutral losses cosine score aims at quantifying the similarity between two mass spectra. The score is calculated by finding best possible matches between peaks of two spectra. Two peaks are considered a potential match if their *m/z* ratios lie within the given `tolerance` once a mass-shift is applied. The mass shift is the difference in precursor-*m/z* between the two spectra.

```{r}
res <- compareSpectriPy(all, caf, param = NeutralLossesCosineParam())
res
```

Note that for the abovementioned similarity calculations all spectra precursor *m/z* values need to be available, otherwise an error will be thrown. Thus, we should always ensure to remove spectra without precursor *m/z* values prior to similarity scoring with this similarity method. Below we remove the precursor *m/z* from one of our input spectra and then show how the `Spectra` object could be subsetted to *valid* spectra for this method.

```{r}
## Remove precursor m/z from the 3rd spectrum
all$precursorMz[3] <- NA

## Filter the input spectra removing those with missing precursor.
all <- all[!is.na(precursorMz(all))]

compareSpectriPy(all, caf, param = ModifiedCosineParam())
```


# Advanced use and internals

For advanced users or developers, the `rspec_to_pyspec()` and
`pyspec_to_rspec()` functions are available that enable conversion between MS
data representations in R and Python (i.e. between the `Spectra::Spectra` object
and the Python `matchms.Spectrum` object). To use these functions the
*reticulate* R package needs to be installed along with a Python environment and
the *matchms* Python package.

To illustrate their use we initialize below the Python environment that is
bundled using the `r BiocStyle::Biocpkg("basilisk")` package within *SpectriPy*.

```{r setup-basilisk}
library(SpectriPy)
library(basilisk)
cl <- basiliskStart(SpectriPy:::matchms_env)
```

We next create a simple `Spectra` object representing fragment spectra for some
small compounds.

```{r}
library(Spectra)

# create example spectra
spd <- DataFrame(
  msLevel = c(2L, 2L, 2L),
  id = c("HMDB0000001", "HMDB0000001", "HMDB0001847"),
  name = c("1-Methylhistidine", "1-Methylhistidine", "Caffeine"))

## Assign m/z and intensity values.
spd$mz <- list(
  c(109.2, 124.2, 124.5, 170.16, 170.52),
  c(83.1, 96.12, 97.14, 109.14, 124.08, 125.1, 170.16),
  c(56.0494, 69.0447, 83.0603, 109.0395, 110.0712,
    111.0551, 123.0429, 138.0662, 195.0876))
spd$intensity <- list(
  c(3.407, 47.494, 3.094, 100.0, 13.240),
  c(6.685, 4.381, 3.022, 16.708, 100.0, 4.565, 40.643),
  c(0.459, 2.585, 2.446, 0.508, 8.968, 0.524, 0.974, 100.0, 40.994))

sps <- Spectra(spd)
```

We next convert the `Spectra` to a `matchms.Spectrum` object.

```{r}
pysps <- rspec_to_pyspec(sps)
pysps
```

As a result we got now a Python list with 3 Spectrum objects containing the peak
data of `sps` as well as a reduced set of the available spectra variables in
`sps`. Which spectra variables get copied to Python can be defined with the
`mapping` parameter of `rspec_to_pyspec`. The default is to convert all
variables defined by `spectraVariableMapping()`, but additional variables along
with their respective names in the Python Spectrum object can be defined
too. Below we list the pre-selected spectra variables that are converted by
default.

```{r}
spectraVariableMapping()
```

With our `Spectra` data converted to Python, we could directly call all routines
from the `matchms` package using the *reticulate* R package. Below we normalize
the intensities of the 3 spectra using the `normalize_intensities()` function
from the *matchms.filtering* library. We thus need to first import the
functionality from that package and can then call the function directly on the
(Python) objects.

```{r}
library(reticulate)
filters <- import("matchms.filtering")

res <- vector("list", length(pysps))
for (i in (seq_along(pysps) - 1))
    res[[i + 1]] <- filters$normalize_intensities(pysps[i])
res <- r_to_py(res)
res
```

We can now convert the list of Python `matchms.Spectrum` objects back to R with
`pyspec_to_rspec()`:

```{r}
sps_r <- pyspec_to_rspec(res)
#' The normalized intensities
intensity(sps_r)
#' The original intensities
intensity(sps)
```

Intensity values have now been normalized to values between 0 and 1.

Note however that, it would be much better (and likely more efficient) to
directly use a Python function on the list of Spectrum objects instead of the
mixed R and Python code used in the example above. Below we define a simple
python script that iterates over the spectra in python and performs the
normalization.

```{r}
py_script <- paste0("from matchms.filtering import normalize_intensities\n",
                    "for i in range(len(pysps)):\n",
                    "    pysps[i] = normalize_intensities(pysps[i])\n")
py$pysps <- pysps
py_run_string(py_script)

tmp <- pyspec_to_rspec(pysps)
intensity(tmp)
```

At very last we need also to stop the Python environment enabled by *basilisk*.

```{r}
basiliskStop(cl)
```

# Session information

```{r}
sessionInfo()
```
