---
title: "Full installation of SpectriPy"
format: html
editor: visual
---

# Introduction

In this document we: - Present a simple way to install SpectriPy

# Installation instructions

System requirements:

-   Install RStudio with latest R version

# Install reticulate

-   Install reticulate for R

```{r}
install.packages("reticulate")
```

# **Install python for use with python3 on reticulate:**

We can use reticulate to install directly python via miniconda A path can be precised to put python *Warning:* If the path for miniconda_reticulate is changed, it needs to be changed in all other steps as well.

```{r}
library(reticulate)
options(timeout = 600)
install_miniconda(path = "~/miniconda3_reticulate",force = TRUE)
```

############################################################################## 

# Set the python environment for reticulate

It is possible to choose which python environment should be used by reticulate. We can here decide to create a specific python environment for reticulate.

Rem: if in step 1 python was created in an other folder, the path to the python3 exe file needs to be specified here in the use_python() function.

A conda environment that can be used in the analysis is also created here.

```{r}
use_python("~/miniconda3_reticulate/bin/python3")
conda_create("r-reticulate")
reticulate::py_config()
```

# Install the python tools needed:

Python packages can now be installed using reticulate in its python environment. To install more python packages, they could be added in this step with the function py_install.

Rem: an error message was printed when "pip = FALSE" for matchms and spectrum_utils. Could be an issue with the env ? On the other hand, pip = FALSE needs to be precised for a correct numpy installation

```{r}
py_install("matchms==0.28.2",pip = TRUE)
py_install("spectrum_utils",pip = TRUE)
py_install("numpy==2.0.2",pip = FALSE)
```

We may also add the python packages to the created conda environment ***WARNING*** python version needs to be specified when installing matchms else it can't be installed here (this needs to be done only when working with a virtual environment)

```{r}
conda_install(envname = "r-reticulate","matchms==0.28.2",pip = TRUE,python_version = "3.12.2")
conda_install(envname = "r-reticulate","spectrum_utils",pip = TRUE)
conda_install(envname = "r-reticulate","numpy==2.0.2",pip = FALSE)
```

# Install all R packages and dependencies

-   If needed, install Bioconductor's *BiocManager* package from CRAN:

    ```{r}
    if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
    BiocManager::install(version = "3.20")
    ```

-   Install all required R/Bioconductor packages: The *option(timeout = 600)* makes sure no error messages are returned if the packages are too long to install

```{r}
options(timeout = 600)
BiocManager::install(c("Spectra", "msdata", "AnnotationHub", "CompoundDb","GenomeInfoDbData","mzR","basilisk"))
```

################################################################################ 

***The installation should now be complete*** we can now try a small example to test it and see if everithing works correctly.

We now needs to restart the R session

## Observations:

reticulate WILL NOT find automatically the python installation. The binary needs to be indicated with use_python() or use_condaenv() as in the example below.

# Example (from the other quarto document)

-   load mzML file (DIA).
-   clean spectra.

***WARNING*** If we DO NOT use the conda environment created (for whatever reason)

We have to run the following command to add python to the PATH else some packages like numpy will not be recognised

The python binary to use needs to be precised in use_python()

```{r}
Sys.setenv(PATH= paste("~/miniconda3_reticulate/bin",Sys.getenv()["PATH"],sep=";"))
library(reticulate)
use_python("~/miniconda3_reticulate/bin/python3")
```

DON'T run this block in any other case

Load required R packages.

**use_condaenv()** is here telling reticulate where the python binary is, for the conda environment that we will use and which python binary to use. *WARNING* as it may not find python if this is not precised.

```{r}
#| warning: false
#' R MS packages
library(Spectra)
library(msdata)

#' libraries for integration with Python; in interactive mode, use
#' `repl_python()` to start a Python shell to execute the Python code
#' blocks and return to R using `exit`
library(reticulate)
use_condaenv("~/miniconda3_reticulate/envs/r-reticulate/bin/python3")
#' To translate between R and Python MS data classes
library(SpectriPy)

#' To retrieve data from Bioconductor's AnnotationHub
library(AnnotationHub)
```

Load a test data set with LC-MS/MS DDA data of a pesticide mix provided by the *matchms* library.

```{r}
fl <- system.file("TripleTOF-SWATH/PestMix1_DDA.mzML", package = "msdata")
pest_dda <- Spectra(fl)
pest_dda$msLevel |>
    table()
```

Load a public reference from Bioconductor's *AnnotationHub*: MassBank release 2023.11:

```{r}
#' Load MassBank release 2023.11 from AnnotationHub
ah <- AnnotationHub()
mbank <- ah[["AH116166"]]
mbank

#' Get the full data as a `Spectra` object
mbank <- Spectra(mbank)
mbank
```

# 

```{python}
import numpy
import matchms
import spectrum_utils
```

Next we extract the MS2 spectra from the `pest_dda` data set and *translate* the `Spectra` to Python.

```{r}
pest_ms2 <- filterMsLevel(pest_dda, 2L)
system.time(
    pest_ms2_py <- rspec_to_pyspec(pest_ms2)
)
pest_ms2_py
```

The translation is not very efficient at present - so we need to work on that (later). Seems the variable lives in R, so to access it we need to append the `r.` to the variable name (check, is this required? can we have the variable in python?).

```{python}
r.pest_ms2_py

```

Object translation is at present not perfect - as an alternative approach we extract the data from the `Spectra` object as base R data types and process these in Python.

```{r}
pest_ms2 <- filterMsLevel(pest_dda, 2L)
pd <- as.list(peaksData(pest_ms2))
precursor_mz <- precursorMz(pest_ms2)
```

We should be able to access these variables directly from Python - could these be used to create an e.g. `spectrum_utils.MsmsSpectrum` directly in Python? Note: each element in the `list` `pd` is a numerical `matrix` (`array`) with two columns, the first with the *m/z*, the second with the intensity values. Question is also which other *spectra variables* (metadata) is required or supported by *spectrum_utils*.

```{python}
r.pd[0]
r.precursor_mz[0]
```

# General observations

-   Setup of R and python environments might be tricky as it seems to involve manual setup steps (please correct and add information/code how to do it properly/semi-automatically).
-   Does this setup work with dockerized R? *basilisk* might be a better alternative then?
-   Limitations of the approach: *translation* of data classes is not very efficient.