---
title: "SpectriPy MS2DeepScore"
author: "EuBIC hackathon team - Matthias Anagho-Mattanovich"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpectriPy MS2DeepScore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Introduction

This vignette demonstrates how to use **MS2DeepScore** within `SpectriPy`, by using `reticulate` to interact with Python-based spectral similarity scoring methods.

**MS2DeepScore** is a machine-learning model that computes spectral similarities based on learned fragmentation patterns, allowing for **molecular structure predictions** from mass spectrometry (MS/MS) spectra.

---

# Installation

Before running this vignette, all required R dependencies need to be installed. Use the following command:

```{r installation}

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("reticulate")
BiocManager::install(c("ggplot2", "reshape2", "gplots", "Spectra", "MsBackendMgf", "rcdk", "fingerprint", "igraph"))
```

---

# **Setting Up Python Environment**

```{r Setting Up Python Environment, results='hide'}
library(reticulate)

# Specify Python 3.10 (Update path)
#use_python("/Users/krv114/Library/r-miniconda-arm64/bin/python3.10", required = TRUE)
use_python("/home/mdegraeve/miniconda3/bin/python", required = TRUE)

# Configure Python
py_config()
```

We also need Python dependencies. Ensure `ms2deepscore`, `matchms`, and `torch` are installed:

```{r Python package installation, results='hide'}
system("pip install ms2deepscore matchms numpy torch", intern = TRUE)
```

This ensures that a specified **Python version** is used for compatibility with `matchms` and `ms2deepscore`.

And finally, we need to download the MS2DeepScore model and a test dataset. if the download times-out, you can increase the timeout limit by changing the options with the commented-out code below.

```{r download needed files}
#options(timeout = 1000)
download.file("https://zenodo.org/records/13897744/files/ms2deepscore_model.pt?download=1", "download_test.pt")
download.file("https://raw.githubusercontent.com/matchms/ms2deepscore/refs/heads/main/tests/resources/pesticides_processed.mgf", "pesticides.mgf")
```

---

# **Initializing the MS2DeepScore Environment**

We define a function to **initialize the environment**, loading Python modules and patching `torch.load()` to allow full model loading.

```{r Adjust Python settings and imports, results='hide'}
initialize_ms2deepscore <- function(python_env = NULL) {
  if (!is.null(python_env)) {
    use_virtualenv(python_env, required = TRUE)
  }

  # Import Python modules
  torch <- import("torch", convert = FALSE)
  ms2ds <- import("ms2deepscore", convert = FALSE)
  matchms <- import("matchms", convert = FALSE)

  # Patch PyTorch to Allow Full Model Loading
  cat("Patching PyTorch to allow full model loading...\n")
  py_run_string("
import torch

# Patch `torch.load()` globally to set `weights_only=False`
original_load = torch.load
def patched_load(*args, **kwargs):
    if 'weights_only' not in kwargs:
        kwargs['weights_only'] = False  # Force weights_only=False
    return original_load(*args, **kwargs)

torch.load = patched_load
  ")

  # Import matchms and MS2DeepScore
  py_run_string("
from matchms.Pipeline import Pipeline, create_workflow
from matchms.filtering.default_pipelines import DEFAULT_FILTERS
from ms2deepscore import MS2DeepScore
")

  return(list(ms2ds = ms2ds, matchms = matchms))
}
```

---

# **Loading the MS2DeepScore Model**

Once the Python environment is ready, we can **load the pre-trained MS2DeepScore model**.

```{r Loading the MS2DeepScore Model, results='hide'}
load_ms2deepscore_model <- function(model_file, env) {
  ms2ds <- env$ms2ds
  cat("Loading MS2DeepScore model...\n")

  model <- ms2ds$models$load_model(model_file)

  if (is.null(model)) {
    stop("ERROR: Model failed to load.")
  } else {
    cat("Model loaded successfully!\n")
  }

  return(model)
}
```

---

# **Running the MS2DeepScore Pipeline**

Now, we define a function to run **MS2DeepScore** on a set of spectra.

```{r Running the MS2DeepScore Pipeline, results='hide'}
run_ms2deepscore_pipeline <- function(spectrum_path, model, env) {
  matchms <- env$matchms
  ms2ds <- env$ms2ds

  cat("Setting up MS2DeepScore pipeline...\n")

  # Run MS2DeepScore in Python
  py_run_string("
pipeline = Pipeline(create_workflow(
    query_filters=DEFAULT_FILTERS,
    score_computations=[[MS2DeepScore, {'model': r.model}]]
))
report = pipeline.run(r.spectrum_path)
similarity_matrix = pipeline.scores.to_array()
")

  # Retrieve results
  cat("Running pipeline...\n")
  similarity_matrix <- py$similarity_matrix 

  cat("Pipeline run successfully!\n")
  return(similarity_matrix)
}
```

---

# **Running the Full Workflow in R**

Now, let’s run the full **MS2DeepScore workflow** in **R**.

```{r Running the Full Workflow in R, results='hide'}
# Define file paths
model_path <- "/Users/krv114/Desktop/MS/R_python_hackathon/ms2deepscore_model.pt"
spectrum_path <- "/Users/krv114/Desktop/MS/R_python_hackathon/pesticides.mgf"

# Initialize Python environment
env <- initialize_ms2deepscore()

# Load the MS2DeepScore model
model <- load_ms2deepscore_model(model_path, env)

# Run the pipeline and get similarity scores
similarity_scores <- run_ms2deepscore_pipeline(spectrum_path, model, env)
```

---

# **Visualizing the Results**

To better understand MS2DeepScore similarity scores, we can visualize them using a **heatmap**.

```{r Plot similarity heatmap and clustered similarity heatmap, fig.show='hold', results='hide'}
library(ggplot2)
library(reshape2)


# Function to plot MS2DeepScore similarity heatmap
plot_similarity_heatmap <- function(similarity_matrix) {
  # Convert matrix to a long format for ggplot2
  similarity_df <- melt(similarity_matrix)

  p <- ggplot(similarity_df, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0.5) +
    labs(title = "MS2DeepScore\nSimilarity Heatmap", x = "Spectrum Index", y = "Spectrum Index") +
    theme_minimal()
  
  print(p)
}

# Run the function
plot_similarity_heatmap(similarity_scores)


plot_clustered_heatmap <- function(similarity_matrix) {
  # Compute hierarchical clustering
  row_order <- hclust(dist(similarity_matrix))$order
  col_order <- hclust(dist(t(similarity_matrix)))$order

  # Convert to long format and reorder based on clustering
  similarity_df <- melt(as.matrix(similarity_matrix))
  similarity_df$Var1 <- factor(similarity_df$Var1, levels = row_order)
  similarity_df$Var2 <- factor(similarity_df$Var2, levels = col_order)

  ggplot(similarity_df, aes(Var1, Var2, fill = value)) +
    geom_tile() +  # Heatmap effect
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0.5) +
    labs(title = "Clustered MS2DeepScore\nSimilarity") + #, x = "Spectrum Index", y = "Spectrum Index") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())  # Hide axis labels
}

# Run function
plot_clustered_heatmap(similarity_scores)
```

---

# **Conclusion**
This vignette demonstrated how to:
**Install and configure MS2DeepScore in R**  
**Load a pre-trained model**  
**Run a similarity scoring pipeline**  
**Visualize the results using a heatmap**  

By integrating **MS2DeepScore** within `SpectriPy`, we can achieve **accurate MS/MS similarity scoring** directly in R.

---

# **Next Steps**
- Compare **MS2DeepScore similarity** to **Tanimoto similarity** from molecular fingerprints.
- Apply **clustering techniques** to group similar spectra.


```{r Plot similarity distribution among the tested compounds and comparison of similarity scores to Tanimoto scores, fig.show='hold', results='hide'}
# Function to plot similarity score distribution
plot_similarity_distribution <- function(similarity_matrix) {
  similarity_values <- as.vector(similarity_matrix)  # Convert to vector
  similarity_values <- similarity_values[similarity_values != 1]  # Remove self-matches
  
  p <- ggplot(data.frame(similarity = similarity_values), aes(x = similarity)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
    labs(title = "Distribution of\nMS2DeepScore\nSimilarity Scores",
         x = "MS2DeepScore Similarity", y = "Count") +
    theme_minimal()
  print(p)
}

# Run the function
plot_similarity_distribution(similarity_scores)


library(Spectra)
library(MsBackendMgf)

# Function to extract SMILES from MGF with manual backend setup
extract_smiles_from_mgf <- function(mgf_file) {
  # Manually initialize the MGF backend
  backend <- MsBackendMgf()
  backend <- backendInitialize(backend, mgf_file)
  
  # Load MGF data using the correct backend
  spectra <- Spectra(backend)
  
  # Print available metadata columns
  cat("Available metadata columns in Spectra:\n")
  print(colnames(spectraData(spectra)))
  
  # Extract SMILES if available
  if (!"SMILES" %in% colnames(spectraData(spectra))) {
    stop("ERROR: 'SMILES' column not found in metadata. Check column names above.")
  }
  
  smiles_list <- spectraData(spectra)$SMILES
  
  # Remove NA values
  smiles_list <- smiles_list[!is.na(smiles_list)]
  
  return(smiles_list)
}

# Run the function
mgf_file <- "/Users/krv114/Desktop/MS/R_python_hackathon/pesticides.mgf"
smiles_list <- extract_smiles_from_mgf(mgf_file)


library(rcdk)
library(fingerprint)

# Function to Compute Tanimoto Similarity from SMILES
compute_tanimoto_similarity <- function(smiles_list) {
  # Convert SMILES to molecular objects
  mols <- lapply(smiles_list, parse.smiles)
  
  # Generate fingerprints
  fingerprints <- lapply(mols, function(mol) {
    if (!is.null(mol[[1]])) {
      return(get.fingerprint(mol[[1]], type = "standard"))
    } else {
      return(NULL)
    }
  })
  
  # Remove NULL values (invalid SMILES)
  fingerprints <- Filter(Negate(is.null), fingerprints)
  
  # Compute Tanimoto similarity matrix
  similarity_matrix <- fp.sim.matrix(fingerprints, method = "tanimoto")
  
  return(as.matrix(similarity_matrix))  # Convert to standard R matrix
}

# Run Tanimoto Similarity Computation
tanimoto_scores <- compute_tanimoto_similarity(smiles_list)


# Function to plot MS2DeepScore vs. Tanimoto Similarity
plot_score_comparison <- function(ms2_scores, tanimoto_scores) {
  df <- data.frame(
    ms2_score = as.vector(ms2_scores),
    tanimoto = as.vector(tanimoto_scores)
  )
  
  p <- ggplot(df, aes(x = tanimoto, y = ms2_score)) +
    geom_point(alpha = 0.5, color = "darkblue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    labs(title = "MS2DeepScore vs.\nTanimoto Similarity",
         x = "Tanimoto Similarity", y = "MS2DeepScore") +
    theme_minimal()
  print(p)
}

# Run the function (assuming `tanimoto_scores` exists)
plot_score_comparison(similarity_scores, tanimoto_scores)
```


```{r Plot molecular similarity network, fig.show='hold', results='hide'}
# Function to extract NAME from MGF with manual backend setup
extract_names_from_mgf <- function(mgf_file) {
  # Manually initialize the MGF backend
  backend <- MsBackendMgf()
  backend <- backendInitialize(backend, mgf_file)
  # Load MGF data using the correct backend
  spectra <- Spectra(backend)
  # Extract NAME if available
  if (!"NAME" %in% colnames(spectraData(spectra))) {
    stop("ERROR: 'NAME' column not found in metadata. Check column names above.")
  }
  names_list <- spectraData(spectra)$NAME
  # Remove NA values
  names_list <- names_list[!is.na(names_list)]
  return(names_list)
}
# Run the function
mgf_file <- "/Users/krv114/Desktop/MS/R_python_hackathon/pesticides.mgf"
names_list <- extract_names_from_mgf(mgf_file)
# Print extracted NAME
print(names_list)


library(igraph)

# Function to plot molecular similarity network
plot_molecular_network <- function(similarity_matrix, threshold = 0.7) {
  # Create a graph from similarity matrix
  adjacency_matrix <- similarity_matrix >= threshold
  graph <- graph_from_adjacency_matrix(adjacency_matrix, mode = "undirected", diag = FALSE)
  
  # Plot the network
  plot(
    graph, vertex.size = 5, vertex.label = NA, 
    #vertex.label.cex = 0.5,
    edge.color = "gray",
    main = "Molecular Similarity\nNetwork (MS2DeepScore)"
  )
}

similarity_scores_named <- similarity_scores
#rownames(similarity_scores_named) <- names_list
#colnames(similarity_scores_named) <- names_list
# Run the function
plot_molecular_network(similarity_scores_named, threshold = 0.7)
```

---

### **Citation**

If using MS2DeepScore, please cite:  
`Huber, F., van der Burg, S., van der Hooft, J. J., & Ridder, L. (2021). MS2DeepScore: a novel deep learning similarity measure to compare tandem mass spectra. Journal of cheminformatics, 13(1), 84.`

---