---
title: "Enabling integration of Python libraries and R packages for combined mass spectrometry data analysis"
package: SpectriPy
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Enabling integration of Python libraries and R packages for combined mass spectrometry data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Mass Spectrometry, MS, MSMS, Metabolomics, Infrastructure, Quantitative}
  %\VignettePackage{SpectriPy}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{Spectra,BiocStyle,SpectriPy,reticulate}
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
BiocStyle::markdown()
```

**Compiled**: `r date()`

# Introduction

Powerful software libraries for mass spectrometry (MS) data are available in
both Python and R, covering specific and sometimes distinct aspects in the
analysis of proteomics and metabolomics data. R's *reticulate* package converts
basic data types between Python and R and enables a seamless interoperability of
both programming languages. The *SpectriPy* package extends *reticulate*
allowing to translate MS data data structures between Python and R. In addition,
functionality from Python's *matchms* library is directly wrapped into R
functions allowing a seamless integration into R based workflows. *SpectriPy*
thus enables powerful proteomics or metabolomics analysis workflows combining
the strengths of Python and R MS libraries.


# Installation

If the *BiocManager* package is not already available, please install it with
`install.packages("BiocManager")`. To install the development version of
*SpectriPy* from GitHub, please install in addition the *remotes* package with
`install.packages("remotes"). As a system dependency, the package requires a
Python environment to be available.

LLLL Update installation instructions.
and can be installed
with the *BiocManager* R package using the command
`BiocManager::install("RforMassSpectrometry/SpectriPy")`. This will install the
latest version of the package from GitHub.
All required python libraries are installed automatically on demand.
Note that first installation or first invocation of specific functions might
take long because of this installation process.


# Translating data structures between Python and R

The [*reticulate*](https://rstudio.github.io/reticulate/) package enables a
seamless integration of Python with R by translating the core data structures
between the two programming languages and sharing a Python (respectively R)
runtime environment that can be accessed from the other programming languages
including shared variables. The *SpectriPy* package builds on that providing
functionality to translate data structures for mass spectrometry (MS) data
between the two languages. Specifically, the package translates between R's
`Spectra` objects (from the `r BiocStyle::Biocpkg("Spectra")` package) and
`matchms.Spectrum` objects from the
[*matchms*](https://github.com/matchms/matchms) Python library.


## R to Python

In the first examples we translated MS data from R to Python. We first load data
from a mzML data file provided through the `r BiocStyle::Biocpkg("msdata")`
package.

```{r}
#' Loading the data from a mzML file as a `Spectra` object.
library(Spectra)
fl <- system.file("TripleTOF-SWATH", "PestMix1_DDA.mzML", package = "msdata")
s_r <- Spectra(fl)
```

We next restrict the data to MS2 spectra and remove spectra with less than 3
peaks (fragments).

```{r}
#' Restrict to MS level 2 spectra
s_r <- filterMsLevel(s_r, 2)
s_r <- s_r[lengths(s_r) < 3]
s_r
```

```{r}
library(AnnotationHub)
ah <- AnnotationHub(localHub = TRUE)
query(ah, "MassBank")
mb <- ah[["AH107049"]]
mb <- Spectra(mb)
```

```{r}
register(SerialParam())
library(MetaboAnnotation)
m <- matchSpectra(s_r, mb, CompareSpectraParam(ppm = 40))
set.seed(123)
idx <- c(whichTarget(m), 5970, 5971, 5972, 5973, 10279, 15505, 15506,
         sample(seq_along(mb), 51))

mb <- mb[sort(idx)]
tmp <- selectSpectraVariables(
    mb, c("msLevel", "rtime", "acquisitionNum", "precursorMz",
          "precursorIntensity", "precursorCharge", "inchi", "smiles",
          "name"))

library(MsBackendMgf)
sm <- c(spectraVariableMapping(MsBackendMgf()), inchi = "INCHI",
        smiles = "SMILES", name = "NAME")
export(tmp, MsBackendMgf(), file = "/home/jo/tmp/test.mgf", mapping = sm)


library(AnnotationHub)
tmp <- system.file()
```





## Python to R

We next import MS data from a file in MGF format using the *matchms* Python
library.



# Spectra similarity calculations using `matchms`

The *SpectriPy* package provides the `compareSpectriPy()` function that allows
to perform spectra similarity calculations using the scoring functions from the
*matchms* Python package. Below all currently supported scoring functions are
listed along with the *parameter* class that allows selecting and configuring
the algorithm in the `compareSpectriPy()` function. Additional functions will be
added in future.

- [*CosineGreedy*](https://matchms.readthedocs.io/en/latest/api/matchms.similarity.CosineGreedy.html):
  `CosineGreedyParam`.
- [*CosineHungarian*](https://matchms.readthedocs.io/en/latest/api/matchms.similarity.CosineHungarian.html):
  `CosineHungarianParam`.
- [*ModifiedCosineParam*](https://matchms.readthedocs.io/en/latest/api/matchms.similarity.ModifiedCosine.html):
  `ModifiedCosineParam`.

We next create some simple example spectra and subsequently use the
`compareSpectriPy()` function to calculate pairwise similarities between these.

```{r, message = FALSE, eval = FALSE}
library(Spectra)
library(SpectriPy)

## Create a Spectra object with two MS2 spectra for Caffeine.
caf <- DataFrame(
    msLevel = c(2L, 2L),
    name = "Caffeine",
    precursorMz = c(195.0877, 195.0877)
)
caf$intensity <- list(
    c(340.0, 416, 2580, 412),
    c(388.0, 3270, 85, 54, 10111))
caf$mz <- list(
    c(135.0432, 138.0632, 163.0375, 195.0880),
    c(110.0710, 138.0655, 138.1057, 138.1742, 195.0864))
caf <- Spectra(caf)

## Create a Spectra object with two MS2 spectra for 1-Methylhistidine
mhd <- DataFrame(
    msLevel = c(2L, 2L),
    precursorMz = c(170.0924, 170.0924),
    id = c("HMDB0000001", "HMDB0000001"),
    name = c("1-Methylhistidine", "1-Methylhistidine"))
mhd$mz <- list(
    c(109.2, 124.2, 124.5, 170.16, 170.52),
    c(83.1, 96.12, 97.14, 109.14, 124.08, 125.1, 170.16))
mhd$intensity <- list(
    c(3.407, 47.494, 3.094, 100.0, 13.240),
    c(6.685, 4.381, 3.022, 16.708, 100.0, 4.565, 40.643))
mhd <- Spectra(mhd)
```

We first calculate pairwise similarities between all spectra defined above and
those of caffeine using *Spectra*'s built-in `compareSpectra()` function.

```{r, eval = FALSE}
all <- c(caf, mhd)
res_r <- compareSpectra(all, caf)
res_r
```

Thus, `compareSpectra()` returned the pairwise similarity scores (by default
calculated using the normalized dot-product function) between all spectra in
`all` (rows) and all spectra in `caf` (columns). `compareSpectriPy()` works
similar, with the difference that we need to specify and configure the
similarity function (from *matchms*) using a dedicated parameter object. Below
we calculate the similarity using the *CosineGreedy* function changing the
`tolerance` to a value of `0.05` (instead of the default `0.1`).

Executing the following step will automatically download and install conda mini
forge to manage Python package dependencies.

```{r, eval = FALSE}
res <- compareSpectriPy(all, caf, param = CosineGreedyParam(tolerance = 0.05))
res
```

As a result `compareSpectriPy()` returns also a numeric matrix of similarities.
Note also that the first `compareSpectriPy()` call takes usually a little longer
because the Python setup has to be initialized.

Next we use the *ModifiedCosine* algorithm that considers also differences
between the spectra's precursor *m/z* in the calculation.

```{r, eval = FALSE}
res <- compareSpectriPy(all, caf, param = ModifiedCosineParam())
res
```

A third example is the *CosineHungarian* algorithm, also a similarity function
from *matchms*. This algorithm calculates the *cosine similarity score* as with
`CosineGreedyParam`, but using the Hungarian algorithm to find the best matching
peaks between the compared spectra. The algorithm can be configured with
parameters `tolerance`, `mzPower` and `intensityPower` (see parameter description for more details).

```{r, eval = FALSE}
res <- compareSpectriPy(all, caf, param = CosineHungarianParam())
res
```

A fourth example is the *NeutralLossesCosine* algorithm, also a similarity
function from *matchms*. The neutral losses cosine score aims at quantifying the
similarity between two mass spectra. The score is calculated by finding best
possible matches between peaks of two spectra. Two peaks are considered a
potential match if their *m/z* ratios lie within the given `tolerance` once a
mass-shift is applied. The mass shift is the difference in precursor-*m/z*
between the two spectra.

```{r, eval = FALSE}
res <- compareSpectriPy(all, caf, param = NeutralLossesCosineParam())
res
```

Note that for the abovementioned similarity calculations all spectra precursor
*m/z* values need to be available, otherwise an error will be thrown. Thus, we
should always ensure to remove spectra without precursor *m/z* values prior to
similarity scoring with this similarity method. Below we remove the precursor
*m/z* from one of our input spectra and then show how the `Spectra` object could
be subsetted to *valid* spectra for this method.

```{r, eval = FALSE}
## Remove precursor m/z from the 3rd spectrum
all$precursorMz[3] <- NA

## Filter the input spectra removing those with missing precursor.
all <- all[!is.na(precursorMz(all))]

compareSpectriPy(all, caf, param = ModifiedCosineParam())
```


# Advanced use and internals

For advanced users or developers, the `rspec_to_pyspec()` and
`pyspec_to_rspec()` functions are available that enable conversion between MS
data representations in R and Python (i.e. between the `Spectra::Spectra` object
and the Python `matchms.Spectrum` object). To use these functions the
*reticulate* R package needs to be installed along with a Python environment and
the *matchms* Python package.

To illustrate their use we initialize below the Python environment that is
bundled using the `r BiocStyle::Biocpkg("basilisk")` package within *SpectriPy*.

```{r setup-basilisk, eval = FALSE}
library(SpectriPy)
library(basilisk)
cl <- basiliskStart(SpectriPy:::matchms_env)
```

We next create a simple `Spectra` object representing fragment spectra for some
small compounds.

```{r, eval = FALSE}
library(Spectra)

# create example spectra
spd <- DataFrame(
  msLevel = c(2L, 2L, 2L),
  id = c("HMDB0000001", "HMDB0000001", "HMDB0001847"),
  name = c("1-Methylhistidine", "1-Methylhistidine", "Caffeine"))

## Assign m/z and intensity values.
spd$mz <- list(
  c(109.2, 124.2, 124.5, 170.16, 170.52),
  c(83.1, 96.12, 97.14, 109.14, 124.08, 125.1, 170.16),
  c(56.0494, 69.0447, 83.0603, 109.0395, 110.0712,
    111.0551, 123.0429, 138.0662, 195.0876))
spd$intensity <- list(
  c(3.407, 47.494, 3.094, 100.0, 13.240),
  c(6.685, 4.381, 3.022, 16.708, 100.0, 4.565, 40.643),
  c(0.459, 2.585, 2.446, 0.508, 8.968, 0.524, 0.974, 100.0, 40.994))

sps <- Spectra(spd)
```

We next convert the `Spectra` to a `matchms.Spectrum` object.

```{r, eval = FALSE}
pysps <- rspec_to_pyspec(sps)
pysps
```

As a result we got now a Python list with 3 Spectrum objects containing the peak
data of `sps` as well as a reduced set of the available spectra variables in
`sps`. Which spectra variables get copied to Python can be defined with the
`mapping` parameter of `rspec_to_pyspec`. The default is to convert all
variables defined by `spectraVariableMapping()`, but additional variables along
with their respective names in the Python Spectrum object can be defined
too. Below we list the pre-selected spectra variables that are converted by
default.

```{r, eval = FALSE}
spectraVariableMapping()
```

With our `Spectra` data converted to Python, we could directly call all routines
from the `matchms` package using the *reticulate* R package. Below we normalize
the intensities of the 3 spectra using the `normalize_intensities()` function
from the *matchms.filtering* library. We thus need to first import the
functionality from that package and can then call the function directly on the
(Python) objects.

```{r, eval = FALSE}
library(reticulate)
filters <- import("matchms.filtering")

res <- vector("list", length(pysps))
for (i in (seq_along(pysps) - 1))
    res[[i + 1]] <- filters$normalize_intensities(pysps[i])
res <- r_to_py(res)
res
```

We can now convert the list of Python `matchms.Spectrum` objects back to R with
`pyspec_to_rspec()`:

```{r, eval = FALSE}
sps_r <- pyspec_to_rspec(res)
#' The normalized intensities
intensity(sps_r)
#' The original intensities
intensity(sps)
```

Intensity values have now been normalized to values between 0 and 1.

Note however that, it would be much better (and likely more efficient) to
directly use a Python function on the list of Spectrum objects instead of the
mixed R and Python code used in the example above. Below we define a simple
python script that iterates over the spectra in python and performs the
normalization.

```{r, eval = FALSE}
py_script <- paste0("from matchms.filtering import normalize_intensities\n",
                    "for i in range(len(pysps)):\n",
                    "    pysps[i] = normalize_intensities(pysps[i])\n")
py$pysps <- pysps
py_run_string(py_script)

tmp <- pyspec_to_rspec(pysps)
intensity(tmp)
```

At very last we need also to stop the Python environment enabled by *basilisk*.

```{r, eval = FALSE}
basiliskStop(cl)
```

# Session information

```{r}
sessionInfo()
```
